---
format: gfm
editor:
  markdown: 
    wrap: 70
bibliography: referencias.bib
---

# Overview

This repository contains the data and R scripts used in the study:

**Bezerra, W. C. A., & Machado, G. (2026). The three-morph problem:
male trimorphism and alternative reproductive tactics in a Neotropical
arachnid. *Animal Behaviour* 237:123614.
10.1016/j.anbehav.2026.123614.**

The study provides the first evidence of male trimorphism in the
Neotropical harvestman *Longiperna concolor* and investigates the
allometric scaling of contest-related traits across morphs. It further
evaluates whether morphological differentiation is associated with
alternative reproductive tactics.

This repository includes the original and processed datasets, and
fully annotated analysis scripts used to generate all results and
figures presented in the manuscript. The materials are provided to
ensure transparency, reproducibility, and long-term accessibility of
the analyses.

# Repository structure

The repository is organized into the following folders and files:

## Data

### Original data

Contains the raw datasets exported directly from field notebooks and
early spreadsheets from the mark-recapture study conducted by
@zatz2011 over the course of an entire reproductive season, between
October 2008 and April 2009. These files have undergone only minimal
manual corrections to address evident field recording errors during
data transcription, along with basic formatting adjustments to ensure
compatibility with R. No systematic cleaning or analytical processing
has been applied.

-   `original_male_data` A 632 × 4 data frame containing
    individual-level information for males marked during the
    mark–recapture study of @zatz2011.
    -   `ID` Individual male id number (factor).
    -   `DSmm` Dorsal scute length in mm, a proxy for body size
        (numeric).
    -   `F4mm` Length of the fourth right femur in mm, a proxy for
        threat device size (numeric).
    -   `MorphZatz` Morph classification into majors and minors
        according to @zatz2011 (factor).
-   `original_harem_fight` A 631 × 4 data frame containing
    individual-level data on harem possession and territorial
    contests. A male was considered to hold a harem in @zatz2011 if
    its modal recapture location coincided with a stable group of
    females (i.e., a harem) and it was captured at that site at least
    four times.
    -   `ID` Individual male id number (factor).
    -   `MorphZatz` Morph classification into majors and minors
        according to @zatz2011 (factor).
    -   `Harem` Binary variable indicating whether each male was
        considered to have a harem, according to @zatz2011 (integer).
    -   `Fight` Binary information on whether each male was observed
        fighting by @zatz2011 (integer).

### Processed data

Processed datasets ready for analysis, including filtered records,
recoded variables, and intermediate results generated by previous
steps, which are used as inputs for subsequent analyses.

-   `male_data`A 631 × 4 data frame generated by the
    `01_exploratory_analysis` script, containing the same information
    as `original_male_data` after the removal of a single anomalous
    observation.
    -   `ID` Individual male id number (factor).
    -   `DSmm` Dorsal scute length in mm, a proxy for body size
        (numeric).
    -   `F4mm` Length of the fourth right femur in mm, a proxy for
        threat device size (numeric).
    -   `MorphZatz` Morph classification into majors and minors
        according to @zatz2011 (factor).
-   `male_morph_data` A 631 × 8 data frame produced by the
    `02_morph_discrimination` script, containing all variables in the
    `male_data` dataset plus morph classifications obtained using
    different probability thresholds of membership.
    -   `ID` Individual male id number (factor).
    -   `DSmm` Dorsal scute length in mm, a proxy for body size
        (numeric).
    -   `F4mm` Length of the fourth right femur in mm, a proxy for
        threat device size (numeric).
    -   `MorphZatz` Morph classification into majors and minors
        according to @zatz2011 (factor).
    -   `T80` Morph classification into alpha, beta and gamma based on
        the bivariate trimodal mixture model, using an 80%
        probability-of-membership threshold (factor).
    -   `T85` Morph classification into alpha, beta and gamma based on
        the bivariate trimodal mixture model, using an 85%
        probability-of-membership threshold (factor).
    -   `T90` Morph classification into alpha, beta and gamma based on
        the bivariate trimodal mixture model, using an 90%
        probability-of-membership threshold (factor).
    -   `T95` Morph classification into alpha, beta and gamma based on
        the bivariate trimodal mixture model, using an 95%
        probability-of-membership threshold (factor).
-   `membership_probability_data` A 631 × 3 data frame generated by
    the `02_morph_discrimination` script, with individuals in rows and
    morphs in columns. It contains the probability of each male
    belonging to each morph under the bivariate trimodal mixture
    model.
    -   `Alpha` Probability of each individual being classified as an
        alpha (numeric).
    -   `Beta` Probability of each individual being classified as a
        beta (numeric).
    -   `Gamma` Probability of each individual being classified as a
        gamma (numeric).
-   `harem_fight` A 631 × 5 data frame produced by the
    `03_behavioral_data` script, containing individual-level
    information on both the dimorphic classification from @zatz2011
    and the trimorphic classification generated using the 80%
    membership-probability threshold, as well as binary indicators of
    whether each individual was observed holding harems and/or
    engaging in territorial contests.
    -   `ID` Individual male id number (factor).
    -   `MorphZatz` Morph classification into majors and minors
        according to @zatz2011 (factor).
    -   `Harem` Binary variable indicating whether each male was
        considered to have a harem, according to @zatz2011 (factor).
    -   `Fight` Binary information on whether each male was observed
        fighting by @zatz2011 (factor).
    -   `T80` Morph classification into alpha, beta and gamma based on
        the bivariate trimodal mixture model, using an 80%
        probability-of-membership threshold (factor).

## Scripts

-   `01_exploratory_analysis.qmd` Initial data exploration and
    summaries used to guide subsequent steps.
-   `02_morph_discrimination.qmd` Classification of males into morphs
    based on univariate and bivariate mixture models.
-   `03_scaling_and_variation.qmd` Analyses of trait exaggeration,
    body size scaling, and inter-morph variation.
-   `04_behavioral_data.qmd` Reanalysis of male territorial behavior
    originally recorded by @zatz2011.

## Supp

Supplementary material figures.

-   `dorsal_scute_length.png` Dorsal scute length histogram for the
    supplementary material.

# How to reproduce the results

These steps restore the exact package environment (via {renv}) and run
the analysis documents in the intended order, reproducing all figures
and outputs as in the manuscript.

## Prerequisites

-   R (\>= 4.5.0) installed.

## Step-by-Step

1.  Clone this repository:

```{r eval=FALSE}
git clone https://github.com/Tixawee/Lconcolor_Trimorphism.git
```

2.  Open the R project file (`Lconcolor_Trimorphism.Rproj`) in
    RStudio.

3.  Restore the package environment using {renv}:

```{r eval=FALSE}
install.packages("renv")       # if not already installed
renv::restore()                # installs packages from renv.lock
```

4.  Run the analysis scripts in order:
    1.  `scripts/01_exploratory_analysis.qmd`
    2.  `scripts/02_morph_discrimination.qmd`
    3.  `scripts/03_scaling_and_variation.qmd`
    4.  `scripts/04_behavioral_data.qmd`

## Session info

The full R session information used in the original analysis is
available in `session_info.txt`.

# License

The scripts and data in this repository are associated with the
manuscript: **The three-morph problem: male trimorphism and
alternative reproductive tactics in a Neotropical arachnid**.

## Usage restrictions:

All scripts and processed datasets are made available under the terms
of the **Creative Commons Attribution 4.0 International License (CC-BY
4.0)**.

You are free to share and adapt the material, provided that
appropriate credit is given.

## Citation:

If you use any part of this repository, please cite the manuscript:

**Bezerra, W. C. A., & Machado, G. (2026). The three-morph problem:
male trimorphism and alternative reproductive tactics in a Neotropical
arachnid. *Animal Behaviour* 237:123614.**

DOI: 10.1016/j.anbehav.2026.123614

# Contact

For questions, suggestions, or collaboration inquiries, please
contact:

**Wellen C. A. Bezerra**

Email: wellen\@ib.usp.br

# Code Club contributors

This section was created for the Code Club Git/GitHub practice.

Add your *last name, position and the year you started in the
laboratory*.

1.  Glauco Machado (forever)
2.  Daniela
3.  Diego
4.  Miguel
5.  Leticia Padovani Holanda (undergraduate: 2024-present)
6.  Laís
7.  Solimary
8.  Renato
9.  Érika
10. Amanda
11. Guilherme
12. Wellen Cristina Alves Bezerra (Master's: 2023–2025; PhD:
    2025–present)

# Code Club conflict

This section was created for the Code Club Git/GitHub practice.

Each pair should modify the same line in their respective branches.
The names below only identify the pairs. Replace them with your own
*last name, position, and year you started in the laboratory*.

1.  Glauco / Wellen
2.  Daniela / Diego
3.  Miguel / Letícia
4.  Laís / Solimary
5.  Renato / Érika
6.  Amanda / Guilherme

# References
